---
title: "Tooth Growth basic exploratory data analysis"
author: "Tomasz Jaskula"
date: "26 avril 2016"
output: pdf_document
---

## Summary

In this project we are going to analyze the tooth growth data. We will use some confidence intervals and hypothesis tests to compare tooth growth by supp and dose.

## Analysis

### Load libraries

```{r, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
```

### Load the data

```{r}
library(datasets)
data(ToothGrowth)
```

### Exploring the data

Let's explore data size

```{r}
dim(ToothGrowth)
```

Structure of the data:

```{r}
str(ToothGrowth)
```

The column `supp` is a factor containing two levels:

* OJ
* VC

Are there any missing data?

```{r}
sum(is.na(ToothGrowth))
```

There is no missing values.

How the head of the data looks like:

```{r}
head(ToothGrowth)
```

Let's see the teeth growth by the dosage

```{r}
ggplot(ToothGrowth, aes(x=factor(dose), y=len, fill=factor(dose))) +
  geom_boxplot() + 
  facet_grid(.~supp) +
  xlab("Dosage") +   
  ylab("Teeth Growth") +  
  ggtitle("Teeth Growth by supplement type")
```

This analysis show us that the dosage affects the tooth length, the larger the dosage, the longer the tooth.

### Confidence intervals and hypothesis testing

This section tries to answer the question if there is a colleration between the tooth growth and the supplement dosage.

The null hypothesis states that there is no correlation between delivery method and tooth length

#### Dosage as factor

We split the data set into 3 datasets, one for each of the doses. This will allow to run the hypothesis test on all the three subsets.

```{r}
dose_05 <-subset(ToothGrowth, dose %in% c(0.5, 1.0))
dose_1 <-subset(ToothGrowth, dose %in% c(0.5, 2.0))
dose_2 <-subset(ToothGrowth, dose %in% c(1.0, 2.0))
```

Let's run the test on each subset

```{r}
t.test(len~dose, data=dose_05, paired=FALSE)
```

```{r}
t.test(len~dose, data=dose_1, paired=FALSE)
```

```{r}
t.test(len~dose, data=dose_2, paired=FALSE)
```

As we see those confidence intervals allows to reject the null hypothesis because there is a correlation between the tooth lenght and the dosage.

#### Supplement as factor

Let's test the effect of supplement on teeth growth

```{r}
t.test(len ~ supp, paired = F, var.equal = F, data = ToothGrowth)
```

The condifence interval doesn't allow to reject the null hypothesis. The supplement type is not correlated to the tooth growth.

## Conclusion

Increasing the dose has an effect on tooth growth however the supplement type has not. We assume that the poplulations are independent.