---
title: "Analysis of the effects of the weather events on public health and economics accros the United States"
output: html_document
---

## Synopsis

Wheater events may have a big impact on the public health and economics. The goal of this analysis is to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and answer the questions which types of events are most harmful with respect to population health and have the greatest economic consequences. The NOAA database tracks characteristics of major storms and weather events in the United States from 1950 to 2011, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Exploring the data

### Loading the data

The data for this analysis can be downloaded from the following locations :

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [bzip2]
* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r, cache=TRUE, message=FALSE, results='hide', warning=FALSE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

dir.create("data")
download.file(url, destfile = "./data/StormData.csv.bz2", method="curl")

stormdata <- read.csv ("./data/StormData.csv.bz2", header=TRUE, stringsAsFactors=FALSE)
```

Let's explore data size

```{r}
dim(stormdata)
```

There is 902297 observations from 1950 to 2011

Structure of the data:

```{r}
str(stormdata)
```

We would be interested in the following columns :

* `EVTYPE`: wheater event type
* `FATALITIES`: the number of fatalities
* `INJURIES`: the number of injuries
* `PROPDMG`: the property damage estimates as actual dollar amounts
* `PROPDMGEXP`: Alphabetical characters used to signify magnitude. For example "K" for thousands, "M" for millions and "B" for billions
* `CROPDMG`: the crop damage estimates as actual dollar amounts
* `CROPDMGEXP`: Alphabetical characters used to signify magnitude. For example "K" for thousands, "M" for millions and "B" for billions

## Analysing and processing the data

Let's subset the data with the features that are of interest

```{r}
data <- stormdata[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
```

There is no missing values in the subset of selected storm data so we don't need to handle it specificlly.

```{r}
sum(is.na(data))
```

